Language Skills in an Ethnically Segmented Labour market: Estonia 1989 -- 2012

### 2018-07-23 paper/ijm.tex
results done

We limit ourselves to men and women in the main working age, between 25 and
55 years old.  Further, we only analyse
individuals who are either working or actively looking for a job.

The dataset allows us to control for standard personal
characteristics and human capital variables, such as age, education
and family status.  The central variables in this study are the same
as in \citep{Leping2008}:

@@

The survey includes self-reported language skills on languages used at
home, and all other languages they understand.  The respondents are
give some guidance in assessing their skills, they can report either
no knowledge, understanding, speaking, writing, or using the
language at home.  We consolidate this information into a dummy variable, equal
to one if one speaks or writes, or uses that language at
home, and zero otherwise.  Hence we disregard passive understanding
only.
  The reader should keep in mind that while the pre-2000 waves include retrospective income, they do not include retrospective language skills. In this way, the results before
1995 are partly based on extrapolation, assuming individual language
skills did not vary before that date.

@@

The average wages, language skills and other explanatory variables together with the relative frequencies for the period 2000--2012 are presented in \cref{tab:descriptive}. We report the data separately for ethnic nationalities (Estonians and Russians) and language
skills (Estonian, Russian and English).
\fxwarning{OT: can you add variable 'year' in this table?  I know it feels a little weird but is great to show time trends.}
\fxwarning{SKB: What do you mean with time trends? Which value should a ``year'' variable take? Give me an example, since I don't understand what you want. }

@@

	\caption*{\small
				Workplace--related variables are reported for employed individuals,
			other variables for the complete workforce.  Sample is limited to persons
			25-55 year old.\\
			\emph{RUS 0} refers to individuals who cannot speak Russian and
			\emph{RUS 1} to those who can. ;Analogously for   
			English
			(ENG) and
			Estonian (EST).\\
			Primary sector = Agriculture,
			Fishing, Mining \\ Secondary sector = Manufacturing,
			Construction, Electricity \\ Tertiary (service) sector =
			Wholesale, Hotels, Transport, Financial intermediation, Real
			estate \\ Public sector = Public administration, Education,
			Health
		}
\end{sidewaystable}%

@@

We analyse the relationship between labour market outcomes and language
skills by ordinary multivariate regression models.
Several related studies use instrumental variables to correct for
omitted variable bias and measurement errors \citep{Chiswick1995, bleakley+chin2004}.
We refrain from doing that. Our decision is mainly data driven,
there are no suitable instruments in our data,\footnote{We run an IV
	regression using age at immigration to instrument Estonian skills
	for those Russians who were in the country at 1990.  The results are
	not statistically significant at any conventional level.} but it also facilitates
comparison with other studies.  For instance,
\citet{azam+2013EDandCC} and \citet{paolo+tansel2015JofDevStud} only
report OLS results, and, even more, find that the available proxy
variables have little influence on the final results.

We use linear probability models (LPM) to estimate the relationship between
language skills and unemployment, mainly in order to avoid unnecessary parametric
assumptions in the corresponding binary choice models.\footnote{The
	results of corresponding probit models are virtually equal to the LPM results.}
Similarly, we estimate the effect of language skills on wage by
OLS.

We describe the outcome of individual $i$ belonging to group $g$ (either Estonian
or Russian) at year $t$ as
\begin{equation}
	\label{eq:specification}
	\begin{split}                                                       
		y_{it} ^{g} = &\: \alpha^{g}{}' \vec{X}_{it}^{g} + \vec{\beta}^{g}{}' \vec{L}_{it}^{g} +
		\eta_{i}^{g} +
		\upsilon_{t} + \epsilon_{it}^{g},                                                                                                                                                          
	\end{split}
\end{equation}
where $y$ is the outcome (either unemployment indicator
or logarithm of net monthly wage), $\vec{X}$ are the other individual
characteristics; $\vec{L}$ is the vector of
language skill descriptors, $\eta_{i}^{g}$ is the individual-specific effect, $\upsilon_{t}$ are the year dummies and $\epsilon_{it}^{g}$
is the idiosyncratic error term. 
%\fxerror{ot: what's the eta term?  do you do random effects?}
%\fxerror{SKB: Eta should be the individual specific random effect. We cluster the standard errors at the individual level, hence estimate a random-effect model, although not a panel random-effect model.}
%\fxerror{ot: why does $\upsilon$ have index $g$?}
%\fxerror{SKB: Should be an error, probably index $g$ was just added to everything.}
As we use panel data, we cluster the
standard errors on individuals.  When analysing the long-term time
trend, we use the same specification (without year fixed effects) to
estimate separate models for each individual year.

We estimate three models that differ in terms of the number of
control variables. The first specification only includes the language skills
and year dummies.  It describes the ``raw'' effect of language skills
substantial time trends in our data.
The second specification additionally includes individual
characteristics (age, education) and family descriptors (marital
status, indicator for inter-ethnic marriage, having kids up to 17 years
old) and county of residence.  
Finally, the third specification also adds industry, occupation, and other
workplace descriptors.
Note that in case of unemployment probability,
we only estimate the first two models.
\fxnote{SKB: Actually, I tried to estimate Model 3 as well for unemployment probability. The language coefficients become insignificant after controlling for industry and occupation. Hence, Estonian language fluency lowers the unemployment probability only if workplace characteristics are ignored. Since nobody questioned this result during review, we should maybe not mention it now. }

@@

The central results for unemployment regressions are given in
\cref{tab:unemployment_estimation_by_sex_and_ethnic_same_sample}\footnote{We
  restrict both models to have the same number of observations as
  requested by a reviewer. The unrestricted table is in the appendix \cref{tab:unemployment_estimation_by_sex_and_ethnic}.}.
The
table clearly indicates that Estonian language skills are associated
with approximately 5 percentage points lower unemployment for both
Russian men and women. We also see that the effect is not radically
influenced by the individual characteristics--for men it is
virtually unchanged while for women the estimate falls by 30\%, from
6.5 to 4.5 percentage points.

@@

Neither Russian nor English show a comparably strong effect.
Knowledge of Russian is not associated with less unemployment for
Estonian men, and is even related to a slightly higher unemployment
for Estonian women. The latter outcome may be related to omitted
variable bias, Russian knowledge was widespread in the Soviet era and may
not have been closely associated with any favourable characteristics,
but has been increasingly more associated with age  during the period of analysis.
English skills are related to less unemployment both for Estonian men
and women, but the effect is small.
For Russians, however, Estonian language skills matter. Those who are
fluent in that language have approximately 5 percentage point less
unemployment, the effect changes only a little when controlling for
individual characteristics.  In contrast, the effect of English
fluency is close to zero and not statistically significant.  As in
case of Estonian, individual characteristics change the estimates only
a little.

@@

\cref{tab:wage_estimation_by_sex_and_ethnic_same_sample} presents the
central wage regression
results.\footnote{As above, we
  restrict both models to have the same number of observations as
  requested by a reviewer.  The unrestricted table is in the appendix \cref{tab:wage_estimation_by_sex_and_ethnic}.}
The table
confirms the descriptive evidence that wage is positively associated
with fluency of all analyzed languages with English showing the
largest effect.  The only remarkable exception is
Estonian language for Russian men where we fail to find any positive
effect for any of the specifications.  This qualitatively repeats the results of
\citet{Toomet2011}. However, for women we observe a sizeable
positive effect between 6 and 16 percentage points, suggesting that
\citet{Toomet2011} results are not valid for the female labour market.  The effect decreases somewhat
(from 0.167 to 0.114) when including the individual characteristics
into the model, suggesting that the premium
is partly caused by education and potentially also by unobserved
individual performance.  Adding observable workplace characteristics
lowers the effects further to 0.066, suggesting that part of the
language skill premium is collected through workplace mobility.

@@

We also see that knowledge of Russian language is associated with a
substantial income premium of 10\% for both Estonian men and women.
The effect falls rapidly when introducing
individual and workplace controls, suggesting that both ability bias
and workplace mobility play a role.

English language is associated with a very large
wage premium, up to 34 log points (model 1 for
Estonian men). It falls substantially when introducing the
individual characteristics into the model (model 2), additional
workplace characteristics have little effect.  However, in all
specifications it is
significantly more
important than Russian for Estonians. The estimates are rather
similar for Estonian and Russian men but little lower for Russian
women than for Estonian women.

In conclusion, the wage regressions suggest that English is the main
language that matters for men.  The effect of Russian is minor and
that of Estonian non-existent.  For women, all three languages are
associated with higher wage but Russian to a smaller extent.
This is in a stark contrast to the unemployment regressions and we return to
this question in the discussion below.

@@

As Estonian economy went through turbulent times and rapid changes
during the 1990s, additional evidence can be gained by a long-term
trends in the estimated effect size.  Here we look at the yearly estimates from 1992 to 2012.
We report the results for model 2, i.e. models that include year
dummies and individual controls, but not workplace
characteristics.\footnote{The results for models with workplace data (model
  3 for wage)
  look qualitatively similar.  More controls in the model reduces the effect size but does not qualitatively
  change it's temporal pattern.}

@@

Now we turn to ethnic Russians (lower panel in the table).  The association between Estonian skills and income shows no
statistically significant results at all.  This is the most unexpected
result in our analysis and we'll return to this
below in section~\ref{sec:segregation_analysis}.
The impact of Estonian language skills for Russian women too, but
since year 2000, we observe a sizable effect of approximately
10\% for several cases.  However, standard errors are large and we
cannot conclude that the effect is positive for all the age groups.
The positive outcomes for women for the second decade in our analysis
suggests that the previously segregated ``Estonian'' and ``Russian''
economies are slowly becoming a more integrated one where ethnic
Russians, at least women, with good Estonian command get better-paid
jobs than those who cannot speak the language.
In comparison, the estimates for English language show
a downward trend for all age groups for both men and for women.  A likely
cause is the increasing abundance of these skills over time.

>>>>

We limit ourselves to men and women in the main working age, between 25 and
55 years old.  Further, we only analyse
individuals who are either working or actively looking for a job.
The dataset allows us to control for standard personal
characteristics and human capital variables, such as age, education
and family status.  We discuss the most important variables below.

@@

The survey includes self-reported language skills on languages used at
home, and all other languages they understand.  The respondents are
given some guidance in assessing their skills, they can report either
no knowledge, understanding only, speaking but not writing, writing, or using the
language at home.  We consolidate this information into a dummy variable, equal
to one if one speaks or writes, or uses that language at
home, and zero otherwise.  Hence we disregard passive understanding
only.

@@

The average wages, language skills and other explanatory variables together with the relative frequencies for the period 2000--2012 are presented in \cref{tab:descriptive}. We report the data separately for ethnic nationalities (Estonians and Russians) and language
skills (Estonian, Russian and English).

@@

	\caption*{\small
				Workplace--related variables are reported for employed individuals,
			other variables for the complete workforce.  Sample is limited to persons
			25-55 year old.\\
			\emph{RUS 0} refers to individuals who do not speak Russian and
			\emph{RUS 1} to those who do.  Analogously for   
			English
			(ENG) and
			Estonian (EST).\\
			Primary sector = Agriculture,
			Fishing, Mining \\ Secondary sector = Manufacturing,
			Construction, Electricity \\ Tertiary (service) sector =
			Wholesale, Hotels, Transport, Financial intermediation, Real
			estate \\ Public sector = Public administration, Education,
			Health
		}
\end{sidewaystable}%

@@

We analyse the relationship between labour market outcomes and language
skills using the ordinary multivariate regression.
Several related studies use instrumental variables to correct for
omitted variable bias and measurement errors \citep{Chiswick1995, bleakley+chin2004}.
We refrain from doing that. Our decision is mainly data driven,
there are no suitable instruments in our data,\footnote{We run an IV
	regression using age at immigration to instrument Estonian skills
	for those Russians who were in the country at 1990.  The results are
	not statistically significant at any conventional level.} but it also facilitates
comparison with other studies.  For instance,
\citet{azam+2013EDandCC} and \citet{paolo+tansel2015JofDevStud} only
report OLS results, and, even more, find that the available proxy
variables have little influence on the final results.

We use linear probability models (LPM) to estimate the relationship between
language skills and unemployment, mainly in order to avoid unnecessary parametric
assumptions in the corresponding binary choice models.\footnote{The
	results of corresponding probit models are virtually identical to the LPM results.}
Similarly, we estimate the effect of language skills on wage by
OLS.

We describe the outcome (log wage and unemployment status) $y$ of individual $i$ at year $t$ as
\begin{equation}
	\label{eq:specification}
	\begin{split}                                                       
		y_{it} = &\: \vec{\alpha}' \vec{X}_{it} + \vec{\beta}{}' \vec{L}_{it} +
		\eta_{i} +
		\upsilon_{t} + \epsilon_{it},                                                                                                                                                          
	\end{split}
\end{equation}
where $\vec{L}$  is the vector of
language skill descriptors, $\vec{X}$ are the other individual
characteristics, $\eta_{i}$ is the individual-specific effect, $\upsilon_{t}$ are the year dummies and $\epsilon_{it}$
is the idiosyncratic error term. 
As we use panel data, we cluster the
standard errors on individuals.  When analysing the long-term time
trend, we use the same specification (without year fixed effects) to
estimate separate models for each individual year.

We estimate three models that differ in terms of 
control variables. The first specification only includes the language skills
and year dummies.  It describes the ``raw'' effect of language skills,
net of strong time trends in our data.
The second specification additionally includes individual
characteristics (age, education) and family descriptors (marital
status, indicator for inter-ethnic marriage, having kids up to 17 years
old) and county of residence.  
Finally, the third specification also adds industry, occupation, and other
workplace descriptors.
Note that in case of unemployment probability,
we only estimate the first two models.

@@

The central results for unemployment regressions are presented in
\cref{tab:unemployment_estimation_by_sex_and_ethnic_same_sample}\footnote{We
  restrict both models to have the same number of observations as
  requested by a reviewer. The unrestricted results are provided in the Appendix (\Cref{tab:unemployment_estimation_by_sex_and_ethnic}).}.
The
table clearly indicates that Estonian language skills are associated
with approximately 5 percentage points lower unemployment for both
Russian men and women. We also see that the effect is not 
influenced in any major way by the individual characteristics--for men it is
virtually unchanged while for women the estimate falls by 30\%, from
6.5 to 4.5 percentage points.

@@

Neither Russian nor English show a comparably strong effect.
Knowledge of Russian is not associated with less unemployment for
Estonian men, and is even related to a slightly higher unemployment
for Estonian women. The latter outcome may be related to omitted
variable bias, Russian knowledge was widespread in the Soviet era and may
not have been closely associated with favorable unobserved
characteristics, however, it has been increasingly associated with age
through the period of analysis.
English skills are related to less unemployment both for Estonian men
and women, but the effect is small.
For Russians, the most important language is clearly Estonian.  Those who are
fluent in that language are approximately 5 percentage point less likely
unemployed, the effect changes only a little when introducing the
individual controls.  In contrast, the effect of English
is close to zero and not statistically significant.  As in
case of Estonian, individual characteristics change the estimates only
a little.

@@

\cref{tab:wage_estimation_by_sex_and_ethnic_same_sample} presents the
central wage regression
results.\footnote{As above, we
  restrict both models to have the same number of observations as
  requested by a reviewer.  The unrestricted table is in the appendix \cref{tab:wage_estimation_by_sex_and_ethnic}.}
The table
confirms the descriptive evidence that wage is positively associated
with fluency of all analyzed languages with English showing the
largest effect.  The only exception is
Estonian language for Russian men where we fail to find any positive
effect for any of the specifications.  This qualitatively repeats the results of
\citet{Toomet2011}. However, for women we observe a sizeable
positive effect between 6 and 16 percentage points, suggesting that
\citet{Toomet2011} results are not valid for the female labour market.  The effect decreases somewhat
(from 0.167 to 0.114) when including the individual characteristics
into the model, suggesting that the premium
is partly caused by education and potentially also by unobserved
individual performance.  Adding observable workplace characteristics
lowers the effects further to 0.066, suggesting that part of the
language skill premium is realized through workplace selection.

@@

We also see that knowledge of Russian language is associated with a
substantial income premium of 10\% for both Estonian men and women.
Individual and workplace controls lower this figure substantially
implying that both ability bias
and workplace mobility play a role.

English language is associated with a very large
wage premium, up to 34 log points, for all the analyzed groups.  The number falls substantially when introducing the
individual characteristics into the model (model 2), additional
workplace characteristics have less effect.  However, in all
specifications it remains
noticeably more
important than Russian for Estonians. The estimates are rather
similar for Estonian and Russian men but little lower for Russian
women than for Estonian women.

In conclusion, the wage regressions suggest that English is the
language most closely associated with income for men.  The effect of Russian is minor and
that of Estonian non-existent.  For women, all three languages are
associated with higher wage but Russian to a smaller extent.
This is in a stark contrast to the unemployment regressions and we return to
this question in the discussion below.

@@

As Estonian economy went through turbulent times and rapid changes
during the 1990s, additional evidence can be gained by a long-term
trends in the estimated effect size.  Here we look at the yearly estimates from 1992 to 2012.
The reader should keep in mind that while the pre-2000 waves include retrospective income, they do not include retrospective language skills. In this way, the results before
1995 are partly based on extrapolation, assuming individual language
skills did not vary before that date.

We report the results for model 2, i.e. models that include year
dummies and individual controls, but not workplace
characteristics.\footnote{The results for models with workplace data (model
  3 for wage)
  look qualitatively similar.  More controls in the model reduces the effect size but does not qualitatively
  change it's temporal pattern.}

@@

Now we turn to ethnic Russians (lower panel in the table).  The association between Estonian skills and income shows no
statistically significant results at all.  This is the most unexpected
result in our analysis and we'll return to this
below in section~\ref{sec:segregation-analysis}.
The impact of Estonian language skills for Russian women too, but
since year 2000, we observe a sizable effect of approximately
10\% for several cases.  However, standard errors are large and we
cannot conclude that the effect is positive for all the age groups.
The positive outcomes for women for the second decade in our analysis
suggests that the previously segregated ``Estonian'' and ``Russian''
economies are slowly becoming a more integrated one where ethnic
Russians, at least women, with good Estonian command get better-paid
jobs than those who cannot speak the language.
In comparison, the estimates for English language show
a downward trend for all age groups for both men and for women.  A likely
cause is the increasing abundance of these skills over time.
